---
title: "Hands-on Exercise 10"
author: "Oscar Ricardo Moreno Gonzalez"
date: "March 28, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Hands-On Exercise \# 10.

# Getting Started

## Install and Launching R Packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse,scales)
```

## The Data

### Importing the Data

```{r}
coffeechain <- readRDS("data/coffeeChain.rds") 
```

### Data Preparation

Aggregate Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## **Bullet chart in ggplot2**

```{r}
#Products ordered by current value
product_plot <- product %>%
  mutate(Product = reorder(Product, current))

ggplot(product_plot, aes(x = Product, y = current)) +

  #benchmark bands
  geom_col(
    aes(y = max(target) * 1.05),
    fill = "grey90",
    width = 0.8
  ) +
  geom_col(
    aes(y = target),
    fill = "grey75",
    width = 0.8
  ) +
  geom_col(
    aes(y = target * 0.67),
    fill = "grey60",
    width = 0.8
  ) +

  #actual vs current performance bar
  geom_col(
    fill = "grey15",
    width = 0.32
  ) +

  #target marker
  geom_errorbar(
    aes(
      ymin = target,
      ymax = target
    ),
    width = 0.45,
    linewidth = 0.9,
    color = "#D55E00"
  ) +

  coord_flip() +

  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.05))
  ) +

  labs(
    title = "Current Performance vs Target",
    subtitle = "Dark bar shows current value; orange line shows target",
    x = NULL,
    y = NULL
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "grey20"),
    axis.text.x = element_text(color = "grey30"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey35"),
    legend.position = "none"
  )
```

## **Plotting sparklines using ggplot2**

1st we create a monthly sales summary by product for 2013 onward, then stores the result in a new dataset (sales_report):

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

We compute the minimum, maximum and end othe the month sales.

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

And compute the 25 and 75 quantiles.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

Plot the Sparkline

```{r}
ggplot(sales_report, aes(x = Month, y = Sales, group = 1)) +
  facet_grid(Product ~ ., scales = "free_y") +

  geom_ribbon(
    data = quarts,
    aes(ymin = quart1, ymax = quart2),
    fill = "grey85",
    alpha = 0.7
  ) +

  geom_line(linewidth = 0.6, color = "grey20") +

  geom_point(data = mins, color = "#D55E00", size = 2) +
  geom_point(data = maxs, color = "#0072B2", size = 2) +

  geom_text(
    data = mins,
    aes(label = comma(Sales)),
    vjust = -0.8,
    size = 3,
    color = "#D55E00"
  ) +
  geom_text(
    data = maxs,
    aes(label = comma(Sales)),
    vjust = 1.6,
    size = 3,
    color = "#0072B2"
  ) +

  geom_text(
    data = ends,
    aes(label = comma(Sales)),
    hjust = 0,
    nudge_x = 0.3,
    size = 3,
    color = "grey30"
  ) +
  geom_text(
    data = ends,
    aes(label = Product),
    hjust = 0,
    nudge_x = 1.1,
    fontface = "bold",
    size = 3.2,
    color = "grey10"
  ) +

  expand_limits(
    x = max(sales_report$Month) +
      0.25 * diff(range(sales_report$Month))
  ) +

  scale_x_continuous(
    breaks = 1:12,
    labels = month.abb
  ) +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0.05, 0.15))
  ) +

  labs(
    title = "Monthly Sales by Product",
    subtitle = "Red points show monthly min, blue points show monthly max",
    caption = "Shaded band = interquartile range"
  ) +

  theme_tufte(base_size = 11, base_family = "Helvetica") +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey30"),
    plot.caption = element_text(size = 9, color = "grey40"),
    axis.text.x = element_text(size = 9, color = "grey20"),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(10, 50, 10, 10)
  )
```

## **Static Dashboard Design (gt and gtExtras)**

```{r}
product %>%
  gt() %>%

  #column labels
  cols_label(
    Product = "Product",
    current = "Performance"
  ) %>%

  #chart column
  gt_plt_bullet(
    column = current,
    target = target,
    width = 70,
    palette = c("#A6CEE3", "#1F78B4")
  ) %>%

  fmt_number(
    columns = c(current, target),
    decimals = 0,
    sep_mark = ","
  ) %>%

  cols_align(
    align = "center",
    columns = current
  ) %>%

  tab_header(
    title = md("**Product Performance vs Target**"),
    subtitle = "Bullet charts compare current values to targets"
  ) %>%

  cols_width(
    Product ~ px(160),
    current ~ px(260)
  ) %>%

  gt_theme_538() %>%

  tab_options(
    table.font.size = px(13),
    data_row.padding = px(6),
    column_labels.font.weight = "bold"
  )
```

## **Sparklines (gtExtras method)**

### Data Preparation

We take the coffee sales data, keep only 2013, group it by product and month and calculate total monthly sales for each product.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

### Plot the report

```{r}
report %>%
  group_by(Product) %>%
  summarize(
    `Monthly Sales` = list(Sales)
  ) %>%

  gt() %>%

  gt_plt_sparkline(
    column = `Monthly Sales`,
    same_limit = FALSE
  ) %>%

  cols_label(
    Product = "Product",
    `Monthly Sales` = "Monthly Trend"
  ) %>%

  tab_header(
    title = md("**Monthly Sales Trends by Product**"),
    subtitle = "Sparkline view of monthly sales across 2013"
  ) %>%

  cols_width(
    Product ~ px(160),
    `Monthly Sales` ~ px(260)
  ) %>%

  cols_align(
    align = "center",
    columns = `Monthly Sales`
  ) %>%

  gt_theme_538() %>%

  tab_options(
    table.font.size = px(13),
    data_row.padding = px(6),
    column_labels.font.weight = "bold"
  )
```

### **Adding statistics**

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### **Combine it with the data.frame**

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

```{r}
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### **Combining bullet chart and sparklines**

We can combining the bullet chart and sparklines using the steps below:

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

## **Interactive Dashboard - reactable and reactablefmtr methods**

We need to install dataui R package to build an interactive sparklines

```{r}
remotes::install_github("timelyportfolio/dataui")
```

```{r}
library(dataui)
```

```{r}
remotes::install_github("kcuilla/reactablefmtr")
```

```{r}
library(reactable)

```

**Plot the interactive sparklines**

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))

```

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

```{r}
reactable(
  report,
  defaultPageSize = 13,
  bordered = FALSE,
  striped = TRUE,
  highlight = TRUE,
  compact = FALSE,
  defaultColDef = colDef(
    headerStyle = list(
      background = "#F7F7F7",
      color = "#222222",
      fontWeight = "bold",
      borderBottom = "2px solid #D9D9D9",
      fontSize = "14px"
    ),
    style = list(
      padding = "10px 12px",
      fontSize = "13px",
      borderBottom = "1px solid #F0F0F0"
    )
  ),
  columns = list(
    Product = colDef(
      maxWidth = 200,
      name = "Product",
      style = list(
        fontWeight = "600",
        color = "#222222"
      )
    ),
    `Monthly Sales` = colDef(
      name = "Monthly Trend",
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"
        ),
        bandline = "innerquartiles",
        statline = "mean"
      ),
      style = list(
        background = "#FCFCFC"
      )
    )
  ),
  style = list(
    border = "1px solid #E6E6E6",
    borderRadius = "8px",
    overflow = "hidden",
    boxShadow = "0 2px 6px rgba(0,0,0,0.06)"
  )
)
```
