---
title: "Take-Home Exercise 1"
author: "Oscar Ricardo Moreno Gonzalez"
date: "February 12, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Exploring the Drivers of Student Academic Performance: A Visual Analytics Approach

## Overview

**Introduction:** This project applies visual analytics to explore factors associated with student learning outcomes using a large-scale survey dataset from the University of Education, Vietnam National University, Hanoi. The dataset captures students demographic characteristics, study behaviors, time allocation, and perceptions of institutional and social support, alongside their academic performance measured by GPA.

**Objective:** Use visually driven analysis to uncover and communicate key patterns in how individual, social, and institutional factors relate to student outcomes.

The analysis emphasizes exploratory visualization to support reasoning, comparison, and interpretation. Using R, *ggplot2*, and selected visualization extensions, the project identifies a set of key observations that highlight meaningful relationships and distributions within the data, demonstrating how visual analytics can help us understand of educational processes and learning environments.

## Getting Started

### Install and Launch R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse, ggcorrplot, ggridges, scales)
```

| Package | Description |
|------------------------------------|------------------------------------|
| tidyverse | A collection of R packages (including *ggplot2*, *dplyr*, and *tidyr*) used for data cleaning, transformation, and visualization. |
| ggcorrplot | Used to visualize correlation matrices in a graphical format. |
| ggridges | Provides ridgeplots that allow comparison of distributions across multiple groups. |
| scales | Enhances visualizations by formatting axes and legends (percentages, labels, breaks, etc.). |

### Data Preprocessing

The original dataset was provided as an CSV file containing numerically coded demographic, behavioral, and perception variables.

-   Data preparation was conducted in R.

-   All categorical and ordinal variables were explicitly recoded as factors with meaningful labels based on the survey documentation (e.g., GPA levels, Likert-scale perceptions, and time-use categories).

-   Ordinal variables were converted to ordered factors to preserve their inherent ranking.

-   No observations were removed during preprocessing, as the dataset contained complete responses.

-   All transformations and analytical visualizations are fully reproducible within this Quarto document.

```{r}
learning_data <- read_csv("data/database.csv")
```

```{r}
learning_data <- learning_data %>%
  mutate(
    GPA = factor(
      GPA,
      levels = 1:5,
      labels = c("Poor", "Average", "Fair", "Good", "Excellent"),
      ordered = TRUE
    ),

    Gender = factor(
      Gender,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),

    Study_Methods = factor(Study_Methods, ordered = TRUE),
    Quality_Lecturer = factor(Quality_Lecturer, ordered = TRUE),
    TrainingCurriculum = factor(TrainingCurriculum, ordered = TRUE)
  )
```

### **Filtering Relevant Variables**

Out of the 22 variables available in the original dataset, a subset of 15 key academic, behavioral, and demographic variables was selected to address the analytical questions. These fields were chosen because they directly relate to student performance, learning behavior and institutional context. Together, they provide a structured framework to examine how individual effort, environmental factors, and socioeconomic context have an impact on academic outcomes.

| Variable | Description |
|------------------------------------|------------------------------------|
| GPA | Ordinal academic performance category ranging from Poor to Excellent. |
| Gender | Binary variable (Male/Female). |
| Time_Studying | Number of hours spent studying per day. |
| Time_SocialMedia | Daily time allocated to social media usage. |
| Study_Methods | Self-reported effectiveness or quality of study strategies. |
| Quality_Lecturer | Student's perception of teaching effectiveness. |
| Adapt_Learning_Uni | Perception of student's adaptation to university. |
| Study_Methods | Effectiveness and organization of student's study strategies and learning approaches. |
| SupportOf_Uni | Perception of academic and administrative support. |
| SupportOf_Lec | Level of guidance, accessibility, and academic assistance from lecturers. |
| Facilitie_Uni | Adequacy and quality of university facilities. |
| Quality_Lecturer | Perception of teaching effectiveness. |
| TrainingCurriculum | Perceived relevance and structure of the curriculum. |
| Competitive_Class | Perceived level of academic competition in the classroom. |
| InfuenceF_Friends | Perceived influence of peers and friends on academic motivation and performance. |

# Analysis

This analysis investigates the factors associated with student academic performance by progressively moving from broad outcome patterns to more specific behavioral and institutional influences.

The sequence is intentionally structured to guide the reader through a logical analytical journey. It begins with a demographic overview of performance differences, then examines how individual time allocation relates to outcomes, before moving toward study behavior and instructional quality. The final step synthesizes these dimensions to highlight how multiple factors interact simultaneously.

## Key Observations

### 1. GPA Distribution by Gender

```{r}
#| echo: false

plot_gpa_gender <- learning_data %>%
  count(GPA, Gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = GPA, y = prop, fill = Gender)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    size = 3,
    color = "white"
  ) +
  scale_fill_manual(values = c(
    "Male" = "grey55",
    "Female" = "steelblue"
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Overall GPA Distribution by Gender",
    x = "GPA Category",
    y = "Proportion of All Students",
    fill = "Gender"
  ) +
  theme_minimal()
```

::: panel-tabset
### Plot

```{r}
plot_gpa_gender
```

### Code

```{#| echo: false}

plot_gpa_gender <- learning_data %>%
  count(GPA, Gender) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = GPA, y = prop, fill = Gender)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_stack(vjust = 0.5),
    size = 3,
    color = "white"
  ) +
  scale_fill_manual(values = c(
    "Male" = "grey55",
    "Female" = "steelblue"
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Overall GPA Distribution by Gender",
    x = "GPA Category",
    y = "Proportion of All Students",
    fill = "Gender"
  ) +
  theme_minimal()
```
:::

This graph show us the distribution of **GPA categories** segmented by **gender**. Most students cluster within the “Fair” and “Good” categories, with relatively few in the “Poor” or “Excellent” extremes. When segmented, female students show a slightly higher proportional concentration in the upper GPA categories, particularly “Good,” while male students appear somewhat more represented in the lower GPA ranges. 

However, the overall pattern remains similar across both groups, indicating that **gender differences** in academic outcomes are **present but not pronounced**. Visually, the chart suggests modest differences rather than large disparities in performance between genders.

### 2. Study Time vs GPA

```{r}
#| echo: false
plot_study_time <- learning_data %>%
  count(Time_Studying, GPA) %>%
  group_by(Time_Studying) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = GPA, y = Time_Studying, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = scales::percent) +
  labs(
    title = "Study Time vs GPA Distribution",
    x = "GPA Category",
    y = "Average Daily Study Time",
    fill = "Proportion"
  ) +
  theme_minimal()
```
::: {.panel-tabset}
### Plot
```{r}
plot_study_time
```

### Code
```{#| echo: false}
plot_study_time <- learning_data %>%
  count(Time_Studying, GPA) %>%
  group_by(Time_Studying) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = GPA, y = Time_Studying, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = scales::percent) +
  labs(
    title = "Study Time vs GPA Distribution",
    x = "GPA Category",
    y = "Average Daily Study Time",
    fill = "Proportion"
  ) +
  theme_minimal()
```
:::
This heatmap visualization represents the distribution of **GPA** categories across different levels of **daily study** **time**. While higher study time categories show a greater concentration of “Good” and “Excellent” outcomes, the **relationship is not strictly linear**.

Moderate study time also display substantial representation in upper GPA categories, suggesting **diminishing returns beyond a certain limit.** (Students reporting extremely high study time do not dominate the top GPA category) Indicating that time investment alone does not fully determine performance.

### 3. Social Media Usage vs GPA

```{r}
#| echo: false
plot_social_media <- learning_data %>%
  count(Time_SocialMedia, GPA) %>%
  group_by(Time_SocialMedia) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = GPA, y = Time_SocialMedia, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue",
                      labels = scales::percent) +
  labs(
    title = "Social Media Usage vs GPA Distribution",
    x = "GPA Category",
    y = "Daily Social Media Time",
    fill = "Proportion"
  ) +
  theme_minimal()

```
::: {.panel-tabset}
### Plot
```{r}
plot_social_media
```
### Code
```{#| echo: false}
plot_social_media <- learning_data %>%
  count(Time_SocialMedia, GPA) %>%
  group_by(Time_SocialMedia) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = GPA, y = Time_SocialMedia, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue",
                      labels = scales::percent) +
  labs(
    title = "Social Media Usage vs GPA Distribution",
    x = "GPA Category",
    y = "Daily Social Media Time",
    fill = "Proportion"
  ) +
  theme_minimal()
```
:::
This second heatmap illustrates the distribution of **GPA** categories across levels of daily **social media usage**. Higher concentrations of “Good” and “Excellent” GPAs appear among students reporting moderate or lower usage, while heavier usage categories show a relatively larger presence in middle GPA tiers. But relationship is not linear, the pattern suggests that balanced time allocation may be more relevant than absolute restriction.

### 4. Study Methods vs GPA

```{r}
#| echo: false
plot_study_methods <- ggplot(learning_data, aes(x = GPA, y = as.numeric(Study_Methods))) +
  geom_violin(fill = "steelblue", alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  labs(
    title = "Perceived Study Methods by GPA",
    x = "GPA Category",
    y = "Study Methods Rating"
  ) +
  theme_minimal()
```
::: {.panel-tabset}
### Plot
```{r}
plot_study_methods
```
### Code
```{#| echo: false}
plot_study_methods <- ggplot(learning_data, aes(x = GPA, y = as.numeric(Study_Methods))) +
  geom_violin(fill = "steelblue", alpha = 0.7) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  labs(
    title = "Perceived Study Methods by GPA",
    x = "GPA Category",
    y = "Study Methods Rating"
  ) +
  theme_minimal()
```
:::
This visualization examines how students perceived effectiveness of their **study methods** varies across **GPA** categories. A gradual upward shift in median ratings is observed as GPA increases, with higher performing students reporting stronger study strategies. 

The graph supports the interpretation that effective study practices are associated with stronger academic outcomes, while also highlighting that performance difference depends on multiple interacting dimensions rather than a single behavioral variable.

### 5. Lecturer Quality vs GPA

```{r}
#| echo: false
plot_lecturer <-ggplot(learning_data, aes(
  x = as.numeric(Quality_Lecturer),
  y = GPA,
  fill = GPA
)) +
  geom_density_ridges(alpha = 0.7, scale = 1.1) +
  scale_fill_manual(values = c(
    "Poor" = "grey80",
    "Average" = "grey65",
    "Fair" = "grey50",
    "Good" = "steelblue3",
    "Excellent" = "steelblue"
  )) +
  labs(
    title = "Perceived Lecturer Quality Across GPA Levels",
    x = "Lecturer Quality Rating",
    y = "GPA Category"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
::: {.panel-tabset}
### Plot
```{r}
plot_lecturer
```
### Code
```{#| echo: false}
plot_lecturer <-ggplot(learning_data, aes(
  x = as.numeric(Quality_Lecturer),
  y = GPA,
  fill = GPA
)) +
  geom_density_ridges(alpha = 0.7, scale = 1.1) +
  scale_fill_manual(values = c(
    "Poor" = "grey80",
    "Average" = "grey65",
    "Fair" = "grey50",
    "Good" = "steelblue3",
    "Excellent" = "steelblue"
  )) +
  labs(
    title = "Perceived Lecturer Quality Across GPA Levels",
    x = "Lecturer Quality Rating",
    y = "GPA Category"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
:::
The ridgeline plot reveals a gradual upward shift in **perceived lecturer quality** as **GPA** increases. Higher performing students tend to report stronger and more concentrated evaluations of lecturer quality, while lower GPA categories display broader and slightly lower distributions. 

The visual progression across GPA levels supports the interpretation that instructional quality contributes positively to student outcomes, but it operates within a broader ecosystem of academic and personal influences. The pattern highlights lecturer quality as a contributor rather than singular factor in performance differentiation.

## 6. Classroom Competition vs GPA

```{r}
#| echo: false
plot_competition <- ggplot(learning_data, aes(x = GPA, y = as.numeric(Competitive_Class))) +
  geom_boxplot(outlier.shape = NA, color = "steelblue") +
  geom_jitter(width = 0.15, alpha = 0.3, color = "steelblue") +
  labs(
    title = "Perceived Classroom Competition by GPA",
    x = "GPA Category",
    y = "Competition Level"
  ) +
  theme_minimal()
```
::: {.panel-tabset}
### Plot
```{r}
plot_competition
```
### Code
```{#| echo: false}
plot_competition <- ggplot(learning_data, aes(x = GPA, y = as.numeric(Competitive_Class))) +
  geom_boxplot(outlier.shape = NA, color = "steelblue") +
  geom_jitter(width = 0.15, alpha = 0.3, color = "steelblue") +
  labs(
    title = "Perceived Classroom Competition by GPA",
    x = "GPA Category",
    y = "Competition Level"
  ) +
  theme_minimal()
```
:::
This plot compares **perceived classroom competition** levels across **GPA** categories. We can see that the distribution of perceived competition levels is relatively consistent across GPA categories, suggesting that classroom competition does not substantially differ between high or low performing students.

We can also notice **variability** is larger among **lower GPA groups**, indicating less consistent perceptions of competition. While moderate competition appears common across all GPA levels, extreme competition levels are less dominant among lower-performing students. The visualization suggests that competition may function as a motivating factor for some students, but its influence likely depends on individual adaptability and study strategies.

### 7. Correlation Matrix of Learning Factors
```{r}
#| echo: false
group_b <- learning_data %>%
  select(
    GPA,
    Adapt_Learning_Uni,
    Study_Methods,
    SupportOf_Uni,
    SupportOf_Lec,
    Facilitie_Uni,
    Quality_Lecturer,
    TrainingCurriculum,
    Competitive_Class,
    InfuenceF_Friends
  ) %>%
  mutate(across(everything(), as.numeric))

corr_matrix <- cor(group_b, use = "complete.obs")

plot_corr <- ggcorrplot(
  corr_matrix,
  type = "lower",
  lab = TRUE,
  lab_size = 3,
  colors = c("white", "grey80", "steelblue")
) +
  labs(title = "Correlation Matrix of Learning Factors and GPA")
```
::: {.panel-tabset}
### Plot
```{r}
plot_corr
```
### Code
```{#| echo: false}
group_b <- learning_data %>%
  select(
    GPA,
    Adapt_Learning_Uni,
    Study_Methods,
    SupportOf_Uni,
    SupportOf_Lec,
    Facilitie_Uni,
    Quality_Lecturer,
    TrainingCurriculum,
    Competitive_Class,
    InfuenceF_Friends
  ) %>%
  mutate(across(everything(), as.numeric))

corr_matrix <- cor(group_b, use = "complete.obs")

plot_corr <- ggcorrplot(
  corr_matrix,
  type = "lower",
  lab = TRUE,
  lab_size = 3,
  colors = c("white", "grey80", "steelblue")
) +
  labs(title = "Correlation Matrix of Learning Factors and GPA")
```
:::
This correlation graph summarizes the relationships among **institutional, behavioral, and social** factors alongside **GPA**. Strong positive correlations are visible among institutional variables such as lecturer quality, university support, and curriculum suitability, suggesting that students perceive these elements as interconnected.

We can see that correlations between these factors and GPA are positive but modest, which indicates that academic performance is influenced by **multiple dimensions** rather than a single dominant variable.

# Conclusion

This project explored how different factors relate to student academic performance. The visuals showed that performance patterns cannot be attributed to a single variable. Instead, they reflect the combined influence of effort, learning strategies, and perceived educational support. The correlation structure show us that these factors interact rather than operate independently.

Overall, the analysis highlights that student success is multidimensional, shaped by both individual practices and the broader learning environment. The findings demonstrate the value of visual exploration in uncovering patterns that may not be immediately evident in raw data and provide a foundation for further investigation into how universities can support effective learning behaviors and academic achievement.
